Comparison of Terrain Following and Cut Cell Grids using a Non-Hydrostatic Model

### 2015-10-21 src/tf-cutcell-comparison/tf-cutcell-comparison.tex
discuss small cell treatment by cubicUpwindCPCFit, tweaks to figures

\TODO{give multiple examples of cut cell grids used in subsequent test cases.  acknowledge that we can still get thin cells in some instances.}

@@

\subsection{Finite volume linear advection model}
The first model discretises the linear advection equation in flux form:
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \mathbf{u} \phi \right) = 0
\label{eq:advect}
\end{align}
where $\mathbf{u} = (u, w)$ is a prescribed velocity field and the tracer density, $\phi$, is interpolated onto cell faces using one of two schemes: first, the centred linear scheme which takes the average of the two neighbouring cell values; second, the upwind-biased multidimensional cubic scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme defined as:
\begin{subequations}
\begin{align}
	\phi^\star &= \phi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\phi^{\star\star} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\phi^{(n+1)} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\mathbf{u} \phi^{(n)})\) at time level \(n\).  This time-stepping scheme is used for consistency with the trapezoidal implicit scheme used for the fully-compressible model, described in section~\ref{sec:model}\ref{sec:euler-model}.
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi(x, z)\), along the edges from \(\Psi\) stored at cell vertices.

@@

\subsection{Fully-compressible model}
\label{sec:euler-model}
The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

Some statistics of the grids are presented in table~\ref{tab:gw-meshes}.  The ratio of minimum and maximum cell areas provides an indication of size of the smallest cut cells.  As expected, there is almost no variation in cell sizes on the BTF grids.  Small cells are generated on cut cell grids at resolutions finer than $\Delta z = \SI{300}{\meter}$ in which the terrain intersects grid layers.  Nonorthogonality is measured by the angle between the vector connecting adjacent cell centres and the normal vector of the shared cell face, ranging from \SI{0}{\degree} on an orthogonal grid, to a maximum of \SI{90}{\degree}.  On the BTF grid, maximum nonorthogonality gradually increases from \SI{9.2}{\degree} to \SI{11.2}{\degree} as increasing horizontal resolution leads to steeper terrain.  Maximum nonorthogonality is higher on almost all cut cell grids.  \TODO{so what?  if we want to talk about nonortho we need a reason for doing so...}

@@

Enlargements of the error field near the mountain is shown in figure~\ref{fig:thermalAdvection}.
\TODO{Errors are negligible on the BTF grids with a maximum error in tracer density of \num{123}.}
Errors are also small on the cut cell grids at $\Delta z = \SI{500}{\meter}$ and $\Delta z = \SI{300}{\meter}$.  However, at resolutions where the terrain intersects layers of the cut cell grids, advection errors are apparent around mountain peaks, with small cells having the largest errors.  These errors are advected horizontally along the lee side, forming stripes.

@@

\TODO{the cubicUpwindCPCFit advection scheme offers an approach for treating small cells that differs from other existing approaches.  the technique will be described in future publication.  (although an implementation of the technique is already available online)  I haven't talked about this anywhere in the body of the paper yet... I need to!}

@@

\begin{table}
	\caption{Cell area ratios and nonorthogonality of BTF and cut cell grids used in the gravity waves and thermal advection tests.  Cell sizes are almost uniform on BTF grids, but for the cut cell grids the cell area ratio gives an indication of the smallest cell sizes.  \TODO{describe nonorthogonality}  \TODO{DOUBLE-CHECK THESE VALUES.}}
	\label{tab:gw-meshes}
%
\centering
\begin{tabular}{S S S S S}
\hline\hline

@@

\begin{figure*}
	\centering
	\includegraphics[width=6in]{../fig-gravityWaves-theta/fig-gravityWaves-theta.pdf}
%
	\caption{Differences in potential temperature between the start and end of the gravity waves test on the BTF grid with $\Delta z = \SI{50}{\meter}$.  Axes are in units of \si{\meter}.  \TODO{draw a dashed line at x=50km if possible}}
	\label{fig:gw-theta}
\end{figure*}

@@

\begin{figure}
	\centering
	\includegraphics{../fig-thermalAdvection/fig-thermalAdvection.pdf}
%
	\caption{Error in tracer density in the thermal advection test at $\Delta z = \SI{50}{\meter}$ on (a) the BTF grid, and (b) the cut cell grid.  \TODO{a sentence of analysis}}
	\label{fig:thermalAdvection}
\end{figure}

>>>>

@@

The two finite volume models make use of the upwind-biased multidimensional cubic advection scheme from \citet{weller-shahrokhi2014} which is non-monotonic and not flux corrected.  
The scheme uses a least squares fit of a multidimensional polynomial using an upwind-biased stencil.  Weightings are iteratively adjusted in the least squares fit procedure to ensure that advection remains upwind-biased where the stencil includes small cut cells.  \TODO{I can see reviewers saying that this isn't a sufficiently detailed explanation...?}

\subsection{Finite volume linear advection model}
The first model discretises the linear advection equation in flux form:
\begin{align}
\partial \phi / \partial t + \nabla \cdot \left( \mathbf{u} \phi \right) = 0
\label{eq:advect}
\end{align}
where $\mathbf{u} = (u, w)$ is a prescribed velocity field and the tracer density, $\phi$, is interpolated onto cell faces using one of two schemes: first, the centred linear scheme which takes the average of the two neighbouring cell values; second, the upwind-biased cubic scheme.
The time derivative is solved using a three-stage, second order Runge-Kutta scheme defined as:
\begin{subequations}
\begin{align}
	\phi^\star &= \phi^{(n)} + \Delta t f(\phi^{(n)}) \\
	\phi^{\star\star} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^\star) \right) \\
	\phi^{(n+1)} &= \phi^{(n)} + \frac{\Delta t}{2} \left( f(\phi^{(n)}) + f(\phi^{\star\star}) \right)
\end{align}
\end{subequations}
where \(f(\phi^{(n)}) = - \nabla \cdot (\mathbf{u} \phi^{(n)})\) at time level \(n\).  This time-stepping scheme is used for consistency with the trapezoidal implicit scheme used for the fully-compressible model, described in section~\ref{sec:model}\ref{sec:euler-model}.
To ensure that the discrete velocity field is non-divergent, velocities are prescribed at cell faces by differencing the streamfunction, \(\Psi(x, z)\), along the edges from \(\Psi\) stored at cell vertices.

@@

\subsection{Finite volume fully-compressible model}
\label{sec:euler-model}
The third model is taken from \citet{weller-shahrokhi2014} which details a discretisation of the fully-compressible Euler equations, given by
\begin{subequations}
\begin{align}
	\text{Momentum} &\ &\  	\frac{\partial \rho \mathbf{u}}{\partial t} + \nabla \cdot \rho \mathbf{u}\otimes\mathbf{u} &= \rho \mathbf{g} - c_p \rho \theta \nabla \Pi - \mu \rho \mathbf{u} \label{eq:momentum} \\
	\text{Continuity} &\ &\	\frac{\partial \rho}{\partial t} + \nabla \cdot \rho \mathbf{u} &= 0 \label{eq:cont} \\
	\text{Thermodynamic equation} &\ &\ \frac{\partial \rho \theta}{\partial t} + \nabla \cdot \rho \mathbf{u} \theta &= 0 \label{eq:theta} \\
	\text{Ideal gas law} &\ &\ \Pi^{(1 - \kappa)/\kappa} &= \frac{R \rho \theta}{p_0} \label{eq:state}
\end{align}
\end{subequations}
where \(\rho\) is the density, \(\mathbf{u}\) is the velocity field, \(\mathbf{g}\) is the gravitational acceleration, \(c_p\) is the heat capacity at constant pressure, \(\theta = T \left(p_0/p\right)^\kappa\) is the potential temperature, \(T\) is the temperature, \(p\) is the pressure, \(p_0\) is a reference pressure, \(\Pi = \left(p / p_0 \right)^\kappa\) is the Exner function of pressure, and \(\kappa = R/c_p\) is the gas constant to heat capacity ratio.  \(\mu\) is a dimensionless damping function used for the sponge layer in the gravity waves test in section~\ref{sec:results}\ref{sec:gw}.

@@

Some statistics of the grids are presented in table~\ref{tab:gw-meshes}.  The ratio of minimum and maximum cell areas provides an indication of size of the smallest cut cells.  As expected, there is almost no variation in cell sizes on the BTF grids.  Small cells are generated on cut cell grids at resolutions finer than $\Delta z = \SI{300}{\meter}$ in which the terrain intersects grid layers.  Nonorthogonality is measured by the angle between the vector connecting adjacent cell centres and the normal vector of the shared cell face, ranging from \SI{0}{\degree} on an orthogonal grid, to a maximum of \SI{90}{\degree}.  The maximum nonorthogonality is the largest value of any face.  On the BTF grid, maximum nonorthogonality gradually increases from \SI{9.2}{\degree} to \SI{11.2}{\degree} as increasing horizontal resolution leads to steeper terrain.  Maximum nonorthogonality is higher on almost all cut cell grids at about \SI{30}{\degree}.  \TODO{so what?  if we want to talk about nonortho we need a reason for doing so...}

@@

Errors are small on the BTF grids with a maximum error in tracer density of \num{0.0376} at $\Delta z = \SI{500}{\meter}$ and errors are 10 to 100 times smaller on higher resolution BTF grids.
Errors are also small on the cut cell grids at $\Delta z = \SI{500}{\meter}$ and $\Delta z = \SI{300}{\meter}$.  However, at resolutions where the terrain intersects layers of the cut cell grids, advection errors are apparent around mountain peaks, with small cells having the largest errors.  These errors are advected horizontally along the lee side, forming stripes.
Enlargements of the error field near the mountain are shown in figure~\ref{fig:thermalAdvection} at $\Delta z = \SI{50}{\meter}$.

@@

The cubic upwind-biased advection scheme takes an approach for treating small cut cells that differs from other existing approaches by adjusting weightings to ensure that advection remains upwind-biased near small cells.  The technique is implemented in OpenFOAM and will be described in a future publication.

@@

\begin{table}
	\caption{Cell area ratios and nonorthogonality of BTF and cut cell grids used in the gravity waves and thermal advection tests.  Cell sizes are almost uniform on BTF grids, but for the cut cell grids the cell area ratio gives an indication of the smallest cell sizes.  Nonorthogonality is the angle between the surface normal vector and the vector joining the two adjacent cell centres.}
	\label{tab:gw-meshes}
%
\centering
\begin{tabular}{S S S S S}
\hline\hline

@@

\begin{figure*}
	\centering
	\includegraphics[width=6in]{../fig-gravityWaves-theta/fig-gravityWaves-theta.pdf}
%
	\caption{Differences in potential temperature between the start and end of the gravity waves test on the BTF grid with $\Delta z = \SI{50}{\meter}$.  The dashed line at $x = \SI{50}{\kilo\meter}$ marks the position of the vertical profile in figure~\ref{fig:gw-sampleLine}.  Axes are in units of \si{\meter}.}
	\label{fig:gw-theta}
\end{figure*}

@@

\begin{figure}
	\centering
	\includegraphics{../fig-thermalAdvection/fig-thermalAdvection.pdf}
%
	\caption{Error in tracer density in the thermal advection test at a resolution of $\Delta z = \SI{50}{\meter}$ on (a) the BTF grid, and (b) the cut cell grid.  Errors are negligible on the BTF grid, but on the cut cell grid errors are generated near mountainous terrain and are advected horizontally on the lee side.  Axes are in units of \si{\meter}.}
	\label{fig:thermalAdvection}
\end{figure}
